---
title: "Summary of Food Associated Compounds - FooDB/Metabolomics "
date: "`r Sys.Date()`"
params:
  food_file: "default.csv"
---

```{css, echo=FALSE}
h1, h2, h3, h4 {
  text-align: center;
}
```

```{r, setup, echo = F}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = T, echo = F, 
                      fig.align = "center")
options(scipen = 999)
```

```{r}
# load necessary packages 
library(readr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(thematic)
library(tidyr)
library(tibble)
library(plotly)
library(ggpubr)
library(stringr)
```

```{r}
# get data 
hmdb <- read_csv("../../Data/hmdb.csv")
meta <- read_csv(params$food_file) 
```

FooDB is a unique database containing the findings of many LC-MS experiments aimed at finding compounds within different food items. It is important to understand what these food compounds are, and how each food is being characterized. Additionally, for each food item there will be multiple experiments assigned to it. For example, an *Apple* will be associated with compounds from experiments on multiple varieties of apples as well as apple byproducts like apple sauce.

**Disclaimer**: Version of FooDB used is the Pre-release 1.0 and is the FooDB CSV file added April 7, 2020. Version 5.0 of HMDB All Metabolites XML file added November 17, 2021. 

<center>
```{r}
# pi chart of all compounds 
meta %>% 
  select(name, kegg_id) %>% 
  group_by(name) %>% 
  distinct() %>% 
  count(name) -> plot_df 


plot_ly(data = plot_df, 
        labels = ~name, 
        values = ~n, 
        type = "pie", 
        sort = F) %>% 
  layout(title = "Number of Compounds Per Food Item (no Duplicates)")
```
</center>

## Detailed Results {.tabset}

### Food Similarity

```{r}
# Summarize compounds and their group status
compound_summary <- meta %>%
  group_by(kegg_id) %>%
  summarise(
    groups = list(unique(name)),
    groups_count = n_distinct(name),
    .groups = "drop"
  ) %>%
  mutate(
    status = ifelse(groups_count == 1, "unique", "shared")
  )

# Count shared compounds (optional, if needed)
shared_count <- compound_summary %>%
  filter(status == "shared") %>%
  summarise(shared_count = n())

# Count unique compounds for each group
compound_summary %>%
  filter(status == "unique") %>%
  unnest(groups) %>%  # Expand the group list
  count(groups, name = "unique_count") -> unique_compounds 

# plot unique compounds and summary stats 
ggplot(unique_compounds, aes(x = reorder(groups, unique_count), y = unique_count)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Food Item", 
       y = "Number of Compounds Unique to Food Item") + 
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  annotate("text", 
           x = length(unique_compounds$groups)/2, 
           y = max(unique_compounds$unique_count) - median(unique_compounds$unique_count) , 
           label = paste0("Numer of Shared Compounds: ", shared_count$shared_count,
                          "\n\nThe Following Statistics are on the \nNumber of Unique Compounds",
                          "\nMean \u00B1 SD: ", round(mean(unique_compounds$unique_count), 2), 
                              "\u00B1", round(sd(unique_compounds$unique_count), 2), 
                          "\nMedian: ", round(median(unique_compounds$unique_count), 2), 
                          "\nRange: ", round(range(unique_compounds$unique_count)[1], 2), "-", 
                              round(range(unique_compounds$unique_count)[2], 2)
                          )
           )
```

### Taxonomic Classification of Compounds 

Points are colored by the number of compounds found in that taxonomic classification. 

```{r}
meta %>% 
  select(name, kegg_id) %>% 
  group_by(name) %>% 
  distinct() %>% 
  left_join(hmdb, by = "kegg_id", 
            relationship = "many-to-many") -> joined

joined %>% 
  count(name, taxonomy_super_class) %>% 
  ggplot(aes(y=taxonomy_super_class, x=name)) +
    geom_point(aes(color = n), size = 4) + 
    scale_color_gradient(low = "yellow", high = "purple4") +
    theme_bw() + 
    theme(legend.title = element_blank(), 
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
    xlab("") + 
    ylab("Taxonomy Super Class ")

joined %>% 
  count(name, taxonomy_class) %>% 
  ggplot(aes(y=taxonomy_class, x=name)) +
    geom_point(aes(color = n), size = 4) + 
    scale_color_gradient(low = "yellow", high = "purple4") +
    theme_bw() + 
    theme(legend.title = element_blank(),
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + 
    xlab("") + 
    ylab("Taxonomy Class ")
```

### Compound Frequency per Food Item

Here each compound is colored by the number of occurrences it was found in LC-MS experiments related to that food item. 

```{r, fig.width=20, fig.height=60}
# plot number of shared compounds per food item
meta %>% 
  select(name, kegg_id) %>% 
  count(name, kegg_id) %>% 
  group_by(name) %>% 
  mutate(Percentage = round(100*(n/sum(n)),2)) %>% 
  ungroup() %>% 
  mutate(bin = cut_number(Percentage, n = 3)) %>% 
  left_join(hmdb, by = "kegg_id") -> freq_comp

ggplot(freq_comp, aes(x = name, y = kegg_id)) + 
  geom_point(aes(color = bin), size = 4) + 
  theme_bw(base_size = 25) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("") + 
  ylab("KEGG compound ID") + 
  labs(color = "Percentage") + 
  facet_wrap(~taxonomy_super_class, scales = "free", 
             labeller = label_wrap_gen(15))
    
```

## Diet Meta-Analysis 

<center>
Identifying similarity and dissimilarity between food items.
```{r}
# make a function to calculte jaccard similarity between food items
jaccard_similarity <- function(food1, food2){ 
  intersection <- length(intersect(food1, food2))
  union <- length(food1) + length(food2) - intersection
  return(intersection/union)
}

# get list of food items
food_items <- unique(meta$name)

# for each pair of food items calculate jaccard similarity and create a list 
jaccard_score <- c()
for(food1 in food_items){
  for (food2 in food_items) {
    
    # get list of unique compounds for food1
    food1_df <- meta %>%
      select(name, kegg_id) %>%
      filter(name == food1) 
    food1_compounds <- unique(food1_df$kegg_id)
      
    # get list of unique compounds for food2
    food2_df <- meta %>%
      select(name, kegg_id) %>%
      filter(name == food2)
    food2_compounds <- unique(food2_df$kegg_id)
    
    # calculate score and add to list 
    score <- jaccard_similarity(food1_compounds, food2_compounds)
    jaccard_score <- c(jaccard_score, c(score))
  }
}

# create matrix 
jaccard_matrix <- matrix(
  jaccard_score, 
  nrow = length(food_items), 
  ncol = length(food_items), 
)
colnames(jaccard_matrix) <- food_items
rownames(jaccard_matrix) <- food_items

# create a long format for matrix
jaccard_long <- as.data.frame(jaccard_matrix) %>% 
  rownames_to_column(var = "Row") %>% 
  pivot_longer(cols = -Row, names_to = "Column", values_to = "Value")

# visualize matrix
ggplot(jaccard_long, aes(x = Column, y = Row, fill = Value)) +
  geom_tile() + 
  scale_fill_gradient(low = "lightblue", high = "darkblue") + 
  labs(title = "Jaccard Matrix Heatmap", x = " ", y = " ", fill = "Simiarity Score") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), 
        plot.title = element_text(hjust = 0.5)) -> p

plotly::ggplotly(p)
```
</center>