---
title: "Graph Query Results Report"
date: "`r Sys.Date()`"
params:
  patterns: "default.csv"
  reactions: "rxn_default.csv"
---

```{css, echo=FALSE, purl=FALSE}
h1, h2, h3, h4 {
  text-align: center;
}
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = TRUE, echo = FALSE,
                      fig.align = "center")
options(scipen = 999)

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyverse)
  library(igraph)
  library(scales)
  library(ggplot2)
  library(ggraph)
  library(jsonlite)
  library(plotly)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(KEGGREST)
})
```

```{r load-data}
hmdb <- read_csv("../../Data/hmdb.csv")
colnames(hmdb)[1] <- "compound"
patterns  <- read_csv(params$patterns)
reactions <- read_json(params$reactions)
```

```{r helper-functions}
# Build igraph nodes + edges and render ggraph
plot_network <- function(df, color_values) {
  c1 <- df %>%
    dplyr::select(compound1_id, compound1_origin, compound1_assoc_food, compound1_freq) %>%
    setNames(c("compound", "origin", "assoc_food", "freq"))

  c2 <- df %>%
    dplyr::select(compound2_id, compound2_origin, compound2_assoc_food, compound2_freq) %>%
    setNames(c("compound", "origin", "assoc_food", "freq"))

  nodes <- rbind(c1, c2) %>%
    distinct(compound, .keep_all = TRUE) %>%
    mutate(freq = replace_na(freq, 0),
           scaled_freq = rescale(freq, to = c(0, 20)))

  edges <- df %>%
    dplyr::select(compound1_id, compound2_id, reaction, KOs, organisms, abundance) %>%
    mutate(scaled_abundance = rescale(abundance, to = c(0, 1)))

  net <- graph_from_data_frame(d = edges, vertices = nodes, directed = TRUE)

  ggraph(net, layout = "fr") +
    geom_edge_link(aes(width = scaled_abundance),
                   arrow = arrow(length = unit(4, "mm"), type = "closed"),
                   end_cap = circle(1, "mm"),
                   show.legend = FALSE) +
    geom_node_point(aes(size = scaled_freq, color = origin)) +
    scale_color_manual(values = color_values) +
    scale_size(guide = "none") +
    geom_node_label(aes(label = name), repel = TRUE, size = 3) +
    theme_void()
}

# Run KEGG enrichment and plot; returns silently if no pathways found
plot_kegg <- function(df) {
  ko_list <- lapply(str_extract_all(df$KOs, "K\\d{5}"), unlist)
  all_kos  <- unique(unlist(ko_list))

  ekegg <- tryCatch(
    enrichKEGG(gene = all_kos, organism = "ko", keyType = "kegg"),
    error = function(e) {
      message("enrichKEGG failed: ", e$message)
      NULL
    }
  )

  if (is.null(ekegg) || nrow(as.data.frame(ekegg)) == 0) {
    message("No significantly enriched KEGG pathways found.")
    return(invisible(NULL))
  }

  print(barplot(ekegg, showCategory = 10))
  print(dotplot(ekegg))
}

# Compounds x Food plots
plot_compounds_food <- function(df, hmdb) {
  food_compounds <- df %>%
    dplyr::select(compound1_id, compound1_assoc_food, compound1_freq) %>%
    mutate(compound1_assoc_food = gsub("\\[|\\]|'", "", compound1_assoc_food)) %>%
    separate_rows(compound1_assoc_food, sep = ",") %>%
    mutate(compound1_assoc_food = str_trim(compound1_assoc_food))

  food_compounds %>%
    group_by(compound1_assoc_food) %>%
    summarise(n_compounds = n_distinct(compound1_id)) %>%
    ggplot(aes(x = n_compounds, y = reorder(compound1_assoc_food, n_compounds))) +
      geom_bar(stat = "identity") +
      theme_bw() +
      xlab("Number of Compounds") +
      ylab(" ") %>%
    print()

  food_compounds <- food_compounds %>%
    dplyr::rename(compound = compound1_id) %>%
    left_join(hmdb, by = "compound")

  for (tax_level in c("taxonomy_super_class", "taxonomy_class")) {
    food_compounds %>%
      dplyr::select(compound, all_of(tax_level), compound1_assoc_food) %>%
      group_by(compound1_assoc_food, .data[[tax_level]]) %>%
      summarise(n_compounds = n_distinct(compound), .groups = "drop") %>%
      ggplot(aes(x = compound1_assoc_food, y = .data[[tax_level]], fill = n_compounds)) +
        geom_tile() +
        theme_bw() +
        scale_y_discrete(labels = label_wrap(width = 15)) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
        xlab(" ") +
        ylab(gsub("_", " ", str_to_title(tax_level))) +
        labs(fill = "Number of \nCompounds") %>%
      print()
  }
}

# Compounds x Genera plots (returns p2, p3 as plotly)
plot_compounds_genera <- function(df, hmdb) {
  microbe_compounds <- df %>%
    dplyr::select(compound2_id, organisms, abundance) %>%
    mutate(organisms = gsub("\\[|\\]|'", "", organisms)) %>%
    separate_rows(organisms, sep = ",") %>%
    mutate(organisms = str_trim(organisms)) %>%
    separate_wider_delim(organisms, delim = ".", names = c("genus", "species"),
                         too_few = "align_start", too_many = "merge", cols_remove = FALSE)

  microbe_compounds %>%
    group_by(genus) %>%
    summarise(n_compounds = n_distinct(compound2_id)) %>%
    ggplot(aes(x = n_compounds, y = reorder(genus, n_compounds))) +
      geom_bar(stat = "identity") +
      theme_bw() +
      xlab("Number of Compounds") +
      ylab(" ") %>%
    print()

  microbe_compounds <- microbe_compounds %>%
    dplyr::rename(compound = compound2_id) %>%
    left_join(hmdb, by = "compound")

  make_tile <- function(tax_level) {
    microbe_compounds %>%
      dplyr::select(compound, all_of(tax_level), genus) %>%
      group_by(genus, .data[[tax_level]]) %>%
      summarise(n_compounds = n_distinct(compound), .groups = "drop") %>%
      ggplot(aes(y = genus, x = .data[[tax_level]], fill = n_compounds)) +
        geom_tile() +
        theme_bw() +
        scale_y_discrete(labels = label_wrap(width = 15)) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
        ylab(" ") +
        xlab(gsub("_", " ", str_to_title(tax_level))) +
        labs(fill = "Number of \nCompounds")
  }

  plotly::ggplotly(make_tile("taxonomy_super_class"))
  plotly::ggplotly(make_tile("taxonomy_class"))
}

# Reactions x Genera heatmap
plot_reactions_genera <- function(df, reactions) {
  reaction_names <- sapply(df$reaction, function(r) {
    name <- reactions[[r]]$NAME
    if (is.null(name)) "unnamed" else name
  })

  reaction_df <- data.frame(reaction = df$reaction, reaction_name = reaction_names)

  p <- df %>%
    left_join(reaction_df, by = "reaction") %>%
    dplyr::select(reaction, organisms, abundance, reaction_name) %>%
    mutate(organisms = gsub("\\[|\\]|'", "", organisms)) %>%
    separate_rows(organisms, sep = ",") %>%
    mutate(organisms = str_trim(organisms)) %>%
    separate_wider_delim(organisms, delim = ".", names = c("genus", "species"),
                         too_few = "align_start", too_many = "merge", cols_remove = FALSE) %>%
    dplyr::select(genus, abundance, reaction_name) %>%
    group_by(genus, reaction_name) %>%
    summarise(s_abundance = sum(abundance), .groups = "drop") %>%
    ggplot(aes(y = reaction_name, x = genus, fill = s_abundance)) +
      geom_tile() +
      theme_bw() +
      scale_y_discrete(labels = label_wrap(width = 15)) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      xlab(" ") +
      ylab("Reaction") +
      labs(fill = "Abundance")

  plotly::ggplotly(p)
}
```

This report is built on data found in `/graph/graph_results.csv`, generated when patterns of interest are searched for. Three queries are performed to look for three different patterns of potential dietary metabolism:

- **(1) food to microbe:** compounds known to only originate from food are used in a microbial reaction whose products only originate from microbes
- **(2) food to both:** compounds known to only originate from food are used in a microbial reaction whose products can originate from both food and microbes
- **(3) both to both:** compounds that can originate from both food and microbes are used in a microbial reaction whose products can also originate from both

This report is divided into two sections: **(1) Most Restrictive Results** (food to microbe patterns only) and **(2) Least Restrictive Results** (food to both and both to both patterns combined).

```{r split-patterns}
patterns1 <- patterns %>%
  filter(compound1_origin == "food" & compound2_origin == "microbe")

# All patterns that are NOT the most restrictive (food -> microbe)
patterns2 <- patterns %>%
  filter(!(compound1_origin == "food" & compound2_origin == "microbe"))
```

## Most Restrictive Results: Food to Microbe

The results of this query show instances where the reactant used in a reaction could only come from a food item, and the resulting product can only come from a microbe.

## {.tabset}

### Pattern Visualization

*Note:* Some edges might represent the same reaction. For example, if reaction1 is `reactant1 --> product1 + product2`, the query creates two rows (reactant1 → product1 and reactant1 → product2), producing one edge per relationship in the graph.

```{r network-1}
if (nrow(patterns1) > 0) {
  plot_network(patterns1, color_values = c("microbe" = "gray50", "food" = "darkcyan"))
} else {
  message("No patterns found for this query.")
}
```

### KEGG Mapper of KOs

*clusterProfiler* [1] is used to functionally profile the KEGG Orthologs (KOs) involved in these reactions. Two standard outputs are shown:

- **Bar plot:** each bar is a significantly enriched KEGG pathway; bar length = number of KOs annotated to that pathway; lighter color = stronger statistical significance.
- **Dot plot:** adds a gene ratio dimension (x-axis = proportion of KOs in pathway / total KOs); dot size = number of KOs in pathway; lighter color = stronger significance.

```{r kegg-1}
if (nrow(patterns1) > 0) plot_kegg(patterns1)
```

### Compounds x Food

```{r compounds-food-1}
if (nrow(patterns1) > 0) plot_compounds_food(patterns1, hmdb)
```

### Compounds x Genera

```{r compounds-genera-1}
if (nrow(patterns1) > 0) plot_compounds_genera(patterns1, hmdb)
```

### Reactions x Genera

```{r reactions-genera-1}
if (nrow(patterns1) > 0) plot_reactions_genera(patterns1, reactions)
```

## Least Restrictive Results: Food to Both & Both to Both

## {.tabset}

### Pattern Visualization

*Note:* Some edges might represent the same reaction. For example, if reaction1 is `reactant1 --> product1 + product2`, the query creates two rows (reactant1 → product1 and reactant1 → product2), producing one edge per relationship in the graph.

```{r network-2}
if (nrow(patterns2) > 0) {
  plot_network(patterns2, color_values = c("both" = "gray50", "food" = "darkcyan"))
} else {
  message("No patterns found for this query.")
}
```

### KEGG Mapper of KOs

*clusterProfiler* [1] is used to functionally profile the KEGG Orthologs (KOs) involved in these reactions. Two standard outputs are shown:

- **Bar plot:** each bar is a significantly enriched KEGG pathway; bar length = number of KOs annotated to that pathway; lighter color = stronger statistical significance.
- **Dot plot:** adds a gene ratio dimension (x-axis = proportion of KOs in pathway / total KOs); dot size = number of KOs in pathway; lighter color = stronger significance.

```{r kegg-2}
if (nrow(patterns2) > 0) plot_kegg(patterns2)
```

### Compounds x Food

```{r compounds-food-2}
if (nrow(patterns2) > 0) plot_compounds_food(patterns2, hmdb)
```

### Compounds x Genera

```{r compounds-genera-2}
if (nrow(patterns2) > 0) plot_compounds_genera(patterns2, hmdb)
```

### Reactions x Genera

```{r reactions-genera-2}
if (nrow(patterns2) > 0) plot_reactions_genera(patterns2, reactions)
```

##

## References

[1] Yu G, Wang L, Han Y and He Q. clusterProfiler: an R package for comparing biological themes among gene clusters. *OMICS: A Journal of Integrative Biology.* 2012, 16(5):284-287.
