---
title: "Graph Query Results Report with Host Information"
date: "`r Sys.Date()`"
params:
  patterns: "default.csv"
  reactions: "rxn_default.csv"
---

```{css, echo=FALSE, purl=FALSE}
h1, h2, h3, h4 {
  text-align: center;
}
```

```{r setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = TRUE, echo = FALSE,
                      fig.align = "center")
options(scipen = 999)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyverse)
  library(igraph)
  library(scales)
  library(ggplot2)
  library(ggraph)
  library(jsonlite)
  library(plotly)
  library(clusterProfiler)
  library(org.Hs.eg.db)
  library(KEGGREST)
})
```

```{r load-data}
hmdb      <- read_csv("../../Data/hmdb.csv")
colnames(hmdb)[1] <- "compound"
patterns  <- read_csv(params$patterns)
reactions <- read_json(params$reactions)
```

```{r helper-functions}
# ── Parsing helpers ──────────────────────────────────────────────────────────

parse_list_col <- function(x) {
  gsub("\\[|\\]|'", "", x)
}

extract_genus <- function(df, org_col) {
  df %>%
    mutate(organisms = parse_list_col(.data[[org_col]])) %>%
    separate_rows(organisms, sep = ",") %>%
    mutate(organisms = str_trim(organisms)) %>%
    separate_wider_delim(organisms, delim = ".",
                         names = c("genus", "species"),
                         too_few = "align_start",
                         too_many = "merge",
                         cols_remove = FALSE)
}

# ── Network ──────────────────────────────────────────────────────────────────

#' Build and plot a directed 3-node-chain network.
#' Edges are compound1→compound2 (edge1) and compound2→compound3 (edge2).
#' Node size ∝ freq; edge width ∝ m_abundance of the relevant edge.
plot_network_3node <- function(df, color_values) {
  if (nrow(df) == 0) { message("No patterns to plot."); return(invisible(NULL)) }

  # Nodes
  n1 <- df %>% dplyr::select(compound = compound1_id, origin = compound1_origin,
                               freq = compound1_freq)
  n2 <- df %>% dplyr::select(compound = compound2_id, origin = compound2_origin,
                               freq = compound2_freq)
  n3 <- df %>% dplyr::select(compound = compound3_id, origin = compound3_origin,
                               freq = compound3_freq)

  nodes <- bind_rows(n1, n2, n3) %>%
    distinct(compound, .keep_all = TRUE) %>%
    mutate(freq = replace_na(freq, 0),
           scaled_freq = rescale(freq, to = c(2, 8)))

  # Edges
  e1 <- df %>% dplyr::select(from = compound1_id, to = compound2_id,
                               abundance = edge1_m_abundance) %>%
    mutate(hop = "edge1")
  e2 <- df %>% dplyr::select(from = compound2_id, to = compound3_id,
                               abundance = edge2_m_abundance) %>%
    mutate(hop = "edge2")

  edges <- bind_rows(e1, e2) %>%
    mutate(scaled_abundance = rescale(abundance, to = c(0.3, 1.5)))

  # Layout via igraph
  net <- igraph::graph_from_data_frame(d = edges, vertices = nodes, directed = TRUE)
  layout_mat <- igraph::layout_with_fr(net)
  coords <- as.data.frame(layout_mat)
  colnames(coords) <- c("x", "y")
  coords$compound <- igraph::V(net)$name

  nodes <- nodes %>% left_join(coords, by = "compound")

  edge_coords <- edges %>%
    left_join(coords %>% dplyr::select(compound, x, y),
              by = c("from" = "compound")) %>%
    dplyr::rename(x_from = x, y_from = y) %>%
    left_join(coords %>% dplyr::select(compound, x, y),
              by = c("to" = "compound")) %>%
    dplyr::rename(x_to = x, y_to = y)

  ggplot() +
    geom_segment(data = edge_coords,
                 aes(x = x_from, y = y_from, xend = x_to, yend = y_to,
                     linewidth = scaled_abundance, linetype = hop),
                 arrow = arrow(length = unit(4, "mm"), type = "closed"),
                 color = "gray40", show.legend = FALSE) +
    geom_point(data = nodes,
               aes(x = x, y = y, color = origin, size = scaled_freq)) +
    scale_color_manual(values = color_values) +
    scale_size(guide = "none") +
    scale_linewidth(guide = "none") +
    ggrepel::geom_label_repel(data = nodes,
                              aes(x = x, y = y, label = compound),
                              size = 3) +
    theme_void()
}

# ── KEGG enrichment ──────────────────────────────────────────────────────────

plot_kegg <- function(ko_vec) {
  all_kos <- unique(unlist(lapply(str_extract_all(ko_vec, "K\\d{5}"), unlist)))
  if (length(all_kos) == 0) { message("No KOs found."); return(invisible(NULL)) }

  ekegg <- tryCatch(
    enrichKEGG(gene = all_kos, organism = "ko", keyType = "kegg"),
    error = function(e) { message("enrichKEGG failed: ", e$message); NULL }
  )

  if (is.null(ekegg) || nrow(as.data.frame(ekegg)) == 0) {
    message("No significantly enriched KEGG pathways found.")
    return(invisible(NULL))
  }

  print(barplot(ekegg, showCategory = 10))
  print(dotplot(ekegg))
}

# ── Food plots ───────────────────────────────────────────────────────────────

plot_compounds_food <- function(df, hmdb,
                                id_col  = "compound1_id",
                                food_col = "compound1_assoc_food",
                                freq_col = "compound1_freq") {
  food_compounds <- df %>%
    dplyr::select(compound_id  = all_of(id_col),
                  assoc_food   = all_of(food_col),
                  freq         = all_of(freq_col)) %>%
    mutate(assoc_food = parse_list_col(assoc_food)) %>%
    separate_rows(assoc_food, sep = ",") %>%
    mutate(assoc_food = str_trim(assoc_food))

  food_compounds %>%
    group_by(assoc_food) %>%
    summarise(n_compounds = n_distinct(compound_id), .groups = "drop") %>%
    ggplot(aes(x = n_compounds, y = reorder(assoc_food, n_compounds))) +
      geom_bar(stat = "identity") +
      theme_bw() +
      xlab("Number of Compounds") + ylab(" ") %>%
    print()

  fc <- food_compounds %>%
    dplyr::rename(compound = compound_id) %>%
    left_join(hmdb, by = "compound")

  for (tax_level in c("taxonomy_super_class", "taxonomy_class")) {
    fc %>%
      dplyr::select(compound, all_of(tax_level), assoc_food) %>%
      group_by(assoc_food, .data[[tax_level]]) %>%
      summarise(n_compounds = n_distinct(compound), .groups = "drop") %>%
      ggplot(aes(x = assoc_food, y = .data[[tax_level]], fill = n_compounds)) +
        geom_tile() + theme_bw() +
        scale_y_discrete(labels = label_wrap(width = 15)) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
        xlab(" ") +
        ylab(gsub("_", " ", str_to_title(tax_level))) +
        labs(fill = "Number of \nCompounds") %>%
      print()
  }
}

# ── Genera plots ─────────────────────────────────────────────────────────────

plot_compounds_genera <- function(df, hmdb,
                                  id_col  = "compound2_id",
                                  org_col = "edge2_organisms",
                                  abun_col = "edge2_m_abundance") {
  mc <- df %>%
    dplyr::select(compound_id = all_of(id_col),
                  organisms   = all_of(org_col),
                  abundance   = all_of(abun_col)) %>%
    extract_genus("organisms") %>%
    dplyr::rename(compound = compound_id) %>%
    left_join(hmdb, by = "compound")

  mc %>%
    group_by(genus) %>%
    summarise(n_compounds = n_distinct(compound), .groups = "drop") %>%
    ggplot(aes(x = n_compounds, y = reorder(genus, n_compounds))) +
      geom_bar(stat = "identity") +
      theme_bw() + xlab("Number of Compounds") + ylab(" ") %>%
    print()

  make_tile <- function(tax_level) {
    mc %>%
      dplyr::select(compound, all_of(tax_level), genus) %>%
      group_by(genus, .data[[tax_level]]) %>%
      summarise(n_compounds = n_distinct(compound), .groups = "drop") %>%
      ggplot(aes(y = genus, x = .data[[tax_level]], fill = n_compounds)) +
        geom_tile() + theme_bw() +
        scale_y_discrete(labels = label_wrap(width = 15)) +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
        ylab(" ") +
        xlab(gsub("_", " ", str_to_title(tax_level))) +
        labs(fill = "Number of \nCompounds")
  }

  plotly::ggplotly(make_tile("taxonomy_super_class"))
  plotly::ggplotly(make_tile("taxonomy_class"))
}

# ── Reactions × Genera ───────────────────────────────────────────────────────

plot_reactions_genera <- function(df, reactions,
                                  rxn_col  = "edge1_reaction",
                                  org_col  = "edge1_organisms",
                                  abun_col = "edge1_m_abundance") {
  rxn_names <- sapply(df[[rxn_col]], function(r) {
    name <- reactions[[r]]$NAME
    if (is.null(name)) "unnamed" else name
  })

  rxn_df <- data.frame(reaction = df[[rxn_col]], reaction_name = rxn_names)

  p <- df %>%
    mutate(reaction_name = rxn_df$reaction_name,
           abundance = .data[[abun_col]]) %>%
    extract_genus(org_col) %>%
    dplyr::select(genus, abundance, reaction_name) %>%
    group_by(genus, reaction_name) %>%
    summarise(s_abundance = sum(abundance), .groups = "drop") %>%
    ggplot(aes(y = reaction_name, x = genus, fill = s_abundance)) +
      geom_tile() + theme_bw() +
      scale_y_discrete(labels = label_wrap(width = 15)) +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
      xlab(" ") + ylab("Reaction") + labs(fill = "Abundance")

  plotly::ggplotly(p)
}

# ── Host abundance plots ─────────────────────────────────────────────────────

#' Compare microbiome vs host abundance across reactions for a given edge.
plot_abundance_comparison <- function(df,
                                      rxn_col   = "edge2_reaction",
                                      m_abun_col = "edge2_m_abundance",
                                      h_abun_col = "edge2_h_abundance") {
  df %>%
    dplyr::select(reaction      = all_of(rxn_col),
                  microbiome    = all_of(m_abun_col),
                  host          = all_of(h_abun_col)) %>%
    distinct() %>%
    pivot_longer(cols = c(microbiome, host),
                 names_to = "source", values_to = "abundance") %>%
    group_by(reaction, source) %>%
    summarise(total_abundance = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(x = reorder(reaction, total_abundance),
               y = total_abundance, fill = source)) +
      geom_bar(stat = "identity", position = "dodge") +
      coord_flip() +
      scale_fill_manual(values = c("microbiome" = "steelblue", "host" = "coral")) +
      theme_bw() +
      xlab("Reaction") + ylab("Total Abundance") +
      labs(fill = "Source")
}

#' Heatmap of host abundance per compound3 origin category.
plot_host_abundance_by_origin <- function(df,
                                           rxn_col   = "edge2_reaction",
                                           h_abun_col = "edge2_h_abundance",
                                           origin_col = "compound3_origin") {
  df %>%
    dplyr::select(reaction = all_of(rxn_col),
                  h_abun   = all_of(h_abun_col),
                  origin   = all_of(origin_col)) %>%
    group_by(reaction, origin) %>%
    summarise(total_h_abundance = sum(h_abun, na.rm = TRUE), .groups = "drop") %>%
    ggplot(aes(x = origin, y = reaction, fill = total_h_abundance)) +
      geom_tile() + theme_bw() +
      scale_y_discrete(labels = label_wrap(width = 20)) +
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
      xlab("Compound 3 Origin") + ylab("Reaction (Edge 2)") +
      labs(fill = "Host\nAbundance")
}

#' Barplot of compound3 endpoint origin distribution.
plot_endpoint_origins <- function(df) {
  df %>%
    count(compound3_origin) %>%
    ggplot(aes(x = n, y = reorder(compound3_origin, n), fill = compound3_origin)) +
      geom_bar(stat = "identity", show.legend = FALSE) +
      theme_bw() +
      xlab("Number of Patterns") + ylab("Compound 3 Origin")
}

#' Scatter: microbiome abundance (edge1) vs host abundance (edge2) per pathway,
#' coloured by compound3 origin.
plot_microbe_vs_host_scatter <- function(df) {
  df %>%
    dplyr::select(compound1   = compound1_id,
                  compound3   = compound3_id,
                  m_abun      = edge1_m_abundance,
                  h_abun      = edge2_h_abundance,
                  origin      = compound3_origin) %>%
    distinct() %>%
    ggplot(aes(x = m_abun, y = h_abun, color = origin,
               text = paste0("C1: ", compound1, "<br>C3: ", compound3))) +
      geom_point(alpha = 0.6, size = 2) +
      scale_x_log10() + scale_y_log10() +
      theme_bw() +
      xlab("Microbiome Abundance (Edge 1, log)") +
      ylab("Host Abundance (Edge 2, log)") +
      labs(color = "Compound 3 Origin") ->
    p
  plotly::ggplotly(p, tooltip = "text")
}
```

This report extends the standard graph query analysis to include **host-derived endpoint information**. Each row in the underlying data represents a **3-node metabolic chain**: a starting compound (compound 1) is transformed by a microbial reaction (edge 1) to an intermediate compound (compound 2), which is then transformed by a second reaction (edge 2) to a final endpoint compound (compound 3). Both microbiome and host abundance values are recorded for each edge, and compound 3 carries an origin label indicating whether the endpoint is primarily host-derived, microbiome-derived, diet-derived, or shared.

The report is divided into two sections mirroring the original: **(1) Most Restrictive** (compound 1 origin = diet only) and **(2) Least Restrictive** (all other origin combinations).

```{r split-patterns}
# Most restrictive: diet -> microbe (intermediate) -> host (endpoint)
patterns1 <- patterns %>%
  filter(compound1_origin == "diet" &
         compound2_origin == "microbe" &
         compound3_origin == "host")

# Least restrictive: everything that is NOT the most restrictive pattern
patterns2 <- patterns %>%
  filter(!(compound1_origin == "diet" &
           compound2_origin == "microbe" &
           compound3_origin == "host"))
```

---

## Most Restrictive Results: Diet → Microbe → Host

These are the strictest patterns: the starting compound (compound 1) is exclusively dietary, the intermediate compound (compound 2) can only originate from microbes, and the endpoint compound (compound 3) is exclusively host-derived. This represents a fully resolved diet-to-microbe-to-host metabolic chain.

## {.tabset}

### Pattern Visualization

*Note:* Solid lines represent edge 1 (compound 1 → compound 2); dashed lines represent edge 2 (compound 2 → compound 3). Edge width reflects microbiome abundance. Node size reflects frequency. Node color indicates compound origin.

```{r network-1}
if (nrow(patterns1) > 0) {
  plot_network_3node(patterns1,
                     color_values = c("diet"      = "darkcyan",
                                      "microbe"   = "gray50",
                                      "microbediet" = "mediumpurple",
                                      "host"      = "tomato",
                                      "hostmicrobe" = "orange",
                                      "hostdiet"  = "goldenrod",
                                      "all"       = "forestgreen"))
} else {
  message("No patterns found for this query.")
}
```

### KEGG Mapper of KOs

*clusterProfiler* [1] is used to functionally profile KOs across both edges. Two standard outputs are shown:

- **Bar plot:** bar length = number of KOs annotated to each enriched KEGG pathway; lighter color = stronger significance.
- **Dot plot:** x-axis = gene ratio (KOs in pathway / total KOs); dot size = number of KOs; lighter color = stronger significance.

```{r kegg-1}
if (nrow(patterns1) > 0) {
  all_kos <- c(patterns1$edge1_KOs, patterns1$edge2_KOs)
  plot_kegg(all_kos)
}
```

### Compounds x Food (Edge 1 Reactant)

These plots show how the diet-origin starting compounds (compound 1) relate to specific food items and chemical taxonomy classes.

```{r compounds-food-1}
if (nrow(patterns1) > 0) {
  plot_compounds_food(patterns1, hmdb,
                      id_col   = "compound1_id",
                      food_col = "compound1_assoc_food",
                      freq_col = "compound1_freq")
}
```

### Compounds x Genera (Edge 1)

Microbial genera performing the first reaction (compound 1 → compound 2), broken down by the chemical taxonomy of compound 2.

```{r genera-edge1-1}
if (nrow(patterns1) > 0) {
  plot_compounds_genera(patterns1, hmdb,
                        id_col   = "compound2_id",
                        org_col  = "edge1_organisms",
                        abun_col = "edge1_m_abundance")
}
```

### Compounds x Genera (Edge 2)

Microbial genera performing the second reaction (compound 2 → compound 3), broken down by the chemical taxonomy of compound 3.

```{r genera-edge2-1}
if (nrow(patterns1) > 0) {
  plot_compounds_genera(patterns1, hmdb,
                        id_col   = "compound3_id",
                        org_col  = "edge2_organisms",
                        abun_col = "edge2_m_abundance")
}
```

### Reactions x Genera (Edge 1)

```{r rxn-genera-edge1-1}
if (nrow(patterns1) > 0) {
  plot_reactions_genera(patterns1, reactions,
                        rxn_col  = "edge1_reaction",
                        org_col  = "edge1_organisms",
                        abun_col = "edge1_m_abundance")
}
```

### Reactions x Genera (Edge 2)

```{r rxn-genera-edge2-1}
if (nrow(patterns1) > 0) {
  plot_reactions_genera(patterns1, reactions,
                        rxn_col  = "edge2_reaction",
                        org_col  = "edge2_organisms",
                        abun_col = "edge2_m_abundance")
}
```

### Host Analysis

**Endpoint origin distribution** — how often does the metabolic chain terminate in a host, microbiome, diet, or mixed-origin compound?

```{r host-origin-1}
if (nrow(patterns1) > 0) {
  plot_endpoint_origins(patterns1)
}
```

**Microbiome vs host abundance by reaction (edge 2)** — comparing how much activity is attributed to microbiome vs host for the second reaction step.

```{r host-abun-compare-1}
if (nrow(patterns1) > 0) {
  plot_abundance_comparison(patterns1,
                             rxn_col    = "edge2_reaction",
                             m_abun_col = "edge2_m_abundance",
                             h_abun_col = "edge2_h_abundance")
}
```

**Host abundance heatmap by reaction and endpoint origin.**

```{r host-heatmap-1}
if (nrow(patterns1) > 0) {
  plot_host_abundance_by_origin(patterns1,
                                 rxn_col    = "edge2_reaction",
                                 h_abun_col = "edge2_h_abundance",
                                 origin_col = "compound3_origin")
}
```

**Microbiome (edge 1) vs host (edge 2) abundance scatter**, coloured by endpoint origin. Points in the upper right represent pathways with high activity on both sides.

```{r scatter-1}
if (nrow(patterns1) > 0) {
  plot_microbe_vs_host_scatter(patterns1)
}
```

---

## Least Restrictive Results: All Other Pattern Combinations

All patterns that do not meet the strict diet → microbe → host criteria above. This aggregates chains where any of the three nodes has a mixed, shared, or alternative origin (e.g. microbediet, hostmicrobe, all).

## {.tabset}

### Pattern Visualization

```{r network-2}
if (nrow(patterns2) > 0) {
  plot_network_3node(patterns2,
                     color_values = c("diet"        = "darkcyan",
                                      "microbe"     = "gray50",
                                      "microbediet" = "mediumpurple",
                                      "host"        = "tomato",
                                      "hostmicrobe" = "orange",
                                      "hostdiet"    = "goldenrod",
                                      "all"         = "forestgreen"))
} else {
  message("No patterns found for this query.")
}
```

### KEGG Mapper of KOs

```{r kegg-2}
if (nrow(patterns2) > 0) {
  all_kos <- c(patterns2$edge1_KOs, patterns2$edge2_KOs)
  plot_kegg(all_kos)
}
```

### Compounds x Food (Edge 1 Reactant)

```{r compounds-food-2}
if (nrow(patterns2) > 0) {
  plot_compounds_food(patterns2, hmdb,
                      id_col   = "compound1_id",
                      food_col = "compound1_assoc_food",
                      freq_col = "compound1_freq")
}
```

### Compounds x Genera (Edge 1)

```{r genera-edge1-2}
if (nrow(patterns2) > 0) {
  plot_compounds_genera(patterns2, hmdb,
                        id_col   = "compound2_id",
                        org_col  = "edge1_organisms",
                        abun_col = "edge1_m_abundance")
}
```

### Compounds x Genera (Edge 2)

```{r genera-edge2-2}
if (nrow(patterns2) > 0) {
  plot_compounds_genera(patterns2, hmdb,
                        id_col   = "compound3_id",
                        org_col  = "edge2_organisms",
                        abun_col = "edge2_m_abundance")
}
```

### Reactions x Genera (Edge 1)

```{r rxn-genera-edge1-2}
if (nrow(patterns2) > 0) {
  plot_reactions_genera(patterns2, reactions,
                        rxn_col  = "edge1_reaction",
                        org_col  = "edge1_organisms",
                        abun_col = "edge1_m_abundance")
}
```

### Reactions x Genera (Edge 2)

```{r rxn-genera-edge2-2}
if (nrow(patterns2) > 0) {
  plot_reactions_genera(patterns2, reactions,
                        rxn_col  = "edge2_reaction",
                        org_col  = "edge2_organisms",
                        abun_col = "edge2_m_abundance")
}
```

### Host Analysis

```{r host-origin-2}
if (nrow(patterns2) > 0) {
  plot_endpoint_origins(patterns2)
}
```

```{r host-abun-compare-2}
if (nrow(patterns2) > 0) {
  plot_abundance_comparison(patterns2,
                             rxn_col    = "edge2_reaction",
                             m_abun_col = "edge2_m_abundance",
                             h_abun_col = "edge2_h_abundance")
}
```

```{r host-heatmap-2}
if (nrow(patterns2) > 0) {
  plot_host_abundance_by_origin(patterns2,
                                 rxn_col    = "edge2_reaction",
                                 h_abun_col = "edge2_h_abundance",
                                 origin_col = "compound3_origin")
}
```

```{r scatter-2}
if (nrow(patterns2) > 0) {
  plot_microbe_vs_host_scatter(patterns2)
}
```

---

## References

[1] Yu G, Wang L, Han Y and He Q. clusterProfiler: an R package for comparing biological themes among gene clusters. *OMICS: A Journal of Integrative Biology.* 2012, 16(5):284-287.
